Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-21 Paper_Draft2/Draft2.Rmd

Our project builds on Women at the Table’s various de-biasing algorithms used in its [AI & Equality Human Rights Toolbox](https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI) to conduct our own analyses on the COMPAS data set. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.

@@

@@

## Statistical Parity Difference

This metric measures the difference that the majority and protected classes get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

## Disparate Impact Ratio

This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].

## Average Odds Difference

This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].

## Equal Opportunity Difference

This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].

>>>>

Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.

@@

**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].

**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].

**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].

@@
